Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-10-13 ms.tex
Finishing Shane's edits to sections 4 and 5

Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of the radiative transfer equation in spherical geometry and is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. The primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center, and we have made a correction to their incorrect $J=0$ surface boundary condition. This correction allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere, and comes from the introduction of the term $J_{\rm bc}$ which is solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. The continuous expansions are discretized and the Fourier coefficients solved for. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, our solutions do not model areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. The peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$; as such, calculations performed at low $\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. This effect is especially prevalent for larger H gas temperature, as the boundary between the Doppler core and line wing extends further out as $T$ increases.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions provided a sufficiently large $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized the $J=0$ surface boundary condition presented in \cite{1973MNRAS.162...43H} and \cite{1990ApJ...350..216N}. For example, in the top-left panel of Figure 1 from \cite{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum (called $H_0$ here) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Though our calculations were performed with a higher gas temperature of $T=10^4$ K, we show in Figure \ref{fig:sol_mc_tau} that the excess present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$. In \citet{2015MNRAS.449.4336S}, the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent, which our boundary condition treatment corrects for. $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in the previously cited solutions. They are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$. This appears to be the consequence of not enforcing the correct frequency-dependent boundary conditions at the surface of the medium.

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. The cumulative distribution function could be calculated using a rejection method, which would then be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method.

This leads in to the primary application of this work: to develop a computationally efficient random walk model in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}. Another approach with wide application is modified random walk methods, such as those discussed in \citet{2002ApJ...567..922A,2015MNRAS.449.4336S} in which Monte Carlo code acceleration is performed for large optical depths in a uniform plane-parallel medium with no absorption. These calculations would need to be generalized to spherical geometry in order to construct a models of resonant scattering in exoplanet atmospheres as discussed in Section \ref{sec:intro}. The underlying principle is the same, however: an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. Such considerations will be taken in future work while developing this solution into a 1D hydrodynamic simulation using radiation forces from accelerated Monte Carlo \lya transfer.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved with an eigenfunction expansion and solved numerically. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

>>>>

A primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We have made a correction to their incorrect $J=0$ surface boundary condition that allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This introduces a term $J_{\rm bc}$ which is solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. The continuous expansions are discretized and the Fourier coefficients solved for. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, our solutions do not model areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Because the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at low $\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. This effect is especially prevalent for larger H gas temperature, as the boundary between the Doppler core and line wing extends further out as $T$ increases.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions provided a sufficiently large $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized the $J=0$ surface boundary condition presented in \cite{1973MNRAS.162...43H} and \cite{1990ApJ...350..216N}. Several papers have previously compared these analytic models to Monte Carlo and seen discrepancies on the order of this correction. For example, in the top-left panel of Figure 1 from \cite{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum (called $H_0$ here) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Though our calculations were performed with a higher gas temperature of $T=10^4$ K, we show in Figure \ref{fig:sol_mc_tau} that the excess present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$. In \citet{2015MNRAS.449.4336S}, the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent. $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in these solutions. They are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$.

The time-dependence of the problem is solved for a $J=0$ boundary condition to characterize the distribution of photon escape times. Solving the differential equation requires a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. 
% TODO: Shane suggests the following places too much emphasis on the number of spatial terms?
It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. 
%
However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. The cumulative distribution function could be calculated using a rejection method, which would then be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method.

This leads in to the primary application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of the radiative transfer equation in spherical geometry and is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a computationally efficient random walk model in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}; an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. Such considerations will be taken in future work while developing this solution into a 1D hydrodynamic simulation using radiation forces from accelerated Monte Carlo \lya transfer.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.
